---
title: "Quality assessment of _Bombus terrestris_ data sets"
author: "Federico Lopez"
date: '`r Sys.Date()`'
output:
  github_document:
    toc: yes
  pdf_document: 
    fig_caption: yes
    toc: yes
  html_document:
    toc: yes
editor_options: 
  chunk_output_type: console
geometry: margin = 1cm
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache.lazy = FALSE,
  include = FALSE,
  out.height = "\textheight",
  out.width = "\textwidth"
)
```

# Load libraries
```{r}
load_pkgs <- function(pkg) {
  sapply(pkg, require, character.only = TRUE)
}

cran_pkgs <- c(
  "BiocManager", "tidyverse", "GGally", "ggtext", "ggh4x", "ggforce",
  "RColorBrewer", "styler", "patchwork", "here"
)

load_pkgs(cran_pkgs)

bioconductor_pkgs <- c(
  "biomaRt", "rhdf5", "tximport", "DESeq2", "vsn", "hexbin"
)

load_pkgs(bioconductor_pkgs)
```

# Set a custom `ggplot` theme
```{r}
# Set a custom ggplot theme
# devtools::install_github("cttobin/ggthemr")
library(ggthemr)

# Diverging
custom_palette <- c(
  "#9E9E9E", "#88CCEE", "#18913D", "#E9C83A", "#882255", "#CC6677",
  "#332288", "#7B73D1", "#2857BA", "#28A7B6", "#F2CB57", "#F2845C"
)

# ggthemr::colour_plot(custom_palette)

custom_theme <- ggthemr::define_palette(
  swatch = custom_palette,
  gradient = c(lower = "#E9C83A", upper = "#882255")
)

ggthemr::ggthemr(custom_theme)
```

```{r}
# here::here()
# Create directory for results
dir.create(here::here("results", "2022-07-20-gene_expression", "supplementary"),
  recursive = TRUE
)
```

# Associate transcripts to genes
```{r}
mart_bter <- biomaRt::useMart(
  biomart = "metazoa_mart",
  dataset = "bterrestris_eg_gene",
  host = "sep2019-metazoa.ensembl.org"
)

tx2gene_bter <- biomaRt::getBM(
  attributes = c("ensembl_transcript_id", "ensembl_gene_id"),
  mart = mart_bter
)

# tx2gene_bter <- read.csv(file = "bter1.0tx2gene.csv",
#   header = TRUE)
```

# Create vector of nAChR gene identifiers
```{r}
nachrs_bter <- c(
  "LOC100643274", "LOC100643282", "LOC100645032", "LOC100646787",
  "LOC100647301", "LOC100647350", "LOC100647624", "LOC100648987",
  "LOC100649515", "LOC100649612", "LOC100649796"
)
```

# Analysis of castes
## Read table of samples and covariates and `kallisto` results
```{r}
# Set sample identifiers using the names of the kallisto output directories
ids_castes_bter <- dir(here::here(
  "results", "2019-colgan_queen_worker_head", "2020-07-14-kallisto", "results"
))

# Read metadata table
table_castes_bter <- read.csv(
  here::here(
    "metadata_bter",
    "2019-colgan_queen_worker_head_metadata.csv"
  ),
  header = TRUE,
  stringsAsFactors = TRUE
)

table_castes_bter$caste <- relevel(
  table_castes_bter$caste,
  ref = "worker"
)

files_castes_bter <- here::here(
  "results", "2019-colgan_queen_worker_head", "2020-07-14-kallisto", "results",
  table_castes_bter$sample,
  "abundance.h5"
)

names(files_castes_bter) <- table_castes_bter$sample

# Check that sample names match
all(ids_castes_bter %in% table_castes_bter$sample)

# Check that all files exist
all(file.exists(files_castes_bter))
```

## Create a `DESeqDataSet` object
```{r}
# Import kallisto quantifications
txi_castes_bter <- tximport::tximport(files_castes_bter,
  type = "kallisto",
  tx2gene = tx2gene_bter,
  txOut = FALSE
)

# Create a DESeqDataSet object
# The design below includes the batch factors "block" and "room"
dds_castes_bter <- DESeq2::DESeqDataSetFromTximport(txi_castes_bter,
  colData = table_castes_bter,
  design = ~ block + room + caste
)

# Choose the reference level for caste factor
dds_castes_bter$caste <- relevel(dds_castes_bter$caste, ref = "worker")
```

## Pre-filter the low-count genes
```{r}
# Estimate the size factors
dds_castes_bter <- DESeq2::estimateSizeFactors(dds_castes_bter)

# Pre-filter the dataset
dim(dds_castes_bter)
# 10661
# At least 4 samples with a count of 10 or higher
# The number of samples would be set to the smallest group size
keep <- rowSums(DESeq2::counts(dds_castes_bter) >= 10) >= 4
dds_castes_bter <- dds_castes_bter[keep, ]
dim(dds_castes_bter)
# 9139
```

## Data transformation and quality assessment
Transform count data using `DESeq2`'s `vst` function.
```{r}
# Apply the variance stabilising transformation
vsd_castes_bter <- DESeq2::vst(dds_castes_bter, blind = FALSE)

# Effect of transformation on the variance
vsn::meanSdPlot(assay(vsd_castes_bter))

ggsave(
  filename = here::here(
    "results", "2022-07-20-gene_expression",
    "supplementary", "bter_castes_vst_counts_smeanSdPlot.pdf"
  ),
  width = 5,
  height = 4,
  units = "in"
)
```

## Remove batch effects using `limma`'s `removeBatchEffect()` function
```{r}
# Removes any shifts in the log2-scale expression data that can be
# explained by batch
# Here "design" relates to "treatment conditions to be preserved, usually the
# design matrix with all experimental factors other than the batch effects."
mod <- model.matrix(~ 0 + caste, data = table_castes_bter)
mat <- assay(vsd_castes_bter)
mat <- limma::removeBatchEffect(mat,
  batch = vsd_castes_bter$block,
  batch2 = vsd_castes_bter$room,
  design = mod
)
assay(vsd_castes_bter) <- mat
```

## Plot the (VST) transformed counts with batch variation removed
```{r}
# Create a tidy data frame with the transformed counts
vsd_nachrs_castes_bter <- assay(vsd_castes_bter) %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "ens_gene") %>%
  tidyr::gather(key = "sample", value = "vst_count", -ens_gene) %>%
  dplyr::filter(ens_gene %in% nachrs_bter) %>%
  dplyr::filter(ens_gene != "LOC100645032") %>%
  dplyr::mutate(subunit = case_when(
    ens_gene == "LOC100647624" ~ "alpha1",
    ens_gene == "LOC100647301" ~ "alpha2",
    ens_gene == "LOC100647350" ~ "alpha3",
    ens_gene == "LOC100648987" ~ "alpha4e5",
    ens_gene == "LOC100649515" ~ "alpha5",
    ens_gene == "LOC100643274" ~ "alpha6",
    ens_gene == "LOC100649796" ~ "alpha7",
    ens_gene == "LOC100643282" ~ "alpha8",
    ens_gene == "LOC100649612" ~ "beta1",
    ens_gene == "LOC100646787" ~ "beta2"
  )) %>%
  dplyr::mutate(caste = case_when(
    grepl("worker", sample) ~ "worker",
    grepl("queen", sample) ~ "queen"
  )) %>%
  dplyr::mutate_if(is.numeric, round, digits = 4)

vsd_nachrs_castes_bter %>%
  dplyr::mutate(
    caste = stringr::str_to_sentence(caste)
  ) %>%
  ggplot(aes(
    x = caste, y = vst_count, color = caste, fill = caste
  )) +
  geom_boxplot(
    alpha = 0.8, width = 0.4, position = position_dodge(0.8), size = 0.6
  ) +
  geom_jitter(
    shape = 21, fill = "white", size = 1.6, stroke = 0.8,
    position = position_dodge(0.8)
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "plain"),
    plot.subtitle = element_text(hjust = 0.5, size = 14, face = "plain"),
    strip.text = element_text(size = 12, face = "plain", hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(size = 12),
    legend.position = "bottom",
    axis.title = element_text(size = 12),
    axis.text.x = element_blank(),
    panel.spacing.x = unit(1, "mm")
  ) +
  labs(
    x = "Caste",
    y = expression(paste("Log"[2], " (VST) transformed count")) # batch variation removed
  ) +
  facet_wrap(~subunit, ncol = 10) +
  scale_color_brewer(palette = "Set2") +
  scale_fill_brewer(palette = "Set2")

ggsave(
  filename = here::here(
    "results", "2022-07-20-gene_expression",
    "supplementary", "bter_castes_vst_counts_nachrs_boxplots.pdf"
  ),
  width = 8,
  height = 5,
  units = "in"
)
```

## Principal component plot of the samples
```{r}
pca_data_castes_bter <- DESeq2::plotPCA(vsd_castes_bter,
  intgroup = "caste",
  n = 500,
  returnData = TRUE
)

# Customize PCA plot
percent_var_castes_bter <- round(100 * attr(pca_data_castes_bter, "percentVar"))

pca_data_castes_bter %>%
  dplyr::mutate(
    caste = stringr::str_to_sentence(caste)
  ) %>%
  ggplot(aes(x = PC1, y = PC2, color = caste)) +
  geom_point(size = 4) +
  xlab(paste0("PC1 [", percent_var_castes_bter[1], "%]")) +
  ylab(paste0("PC2 [", percent_var_castes_bter[2], "%]")) +
  coord_fixed(ratio = 1) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "plain"),
    plot.subtitle = element_text(hjust = 0.5, size = 14, face = "plain"),
    legend.title = element_blank(),
    legend.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  ) +
  labs(
    title = "Relationships among samples of heads of bumble bees",
    subtitle = "(batch effects removed)"
  ) +
  scale_color_brewer(palette = "Set2") +
  scale_fill_brewer(palette = "Set2")

ggsave(
  filename = here::here(
    "results", "2022-07-20-gene_expression",
    "supplementary", "bter_castes_vst_counts_pca.pdf"
  ),
  width = 6,
  height = 6,
  units = "in"
)
```

# Analysis of life stages
## Read `kallisto` results and table of samples and covariates.
```{r}
# Set sample identifiers using the names of the kallisto output directories
ids_lifestages_bter <- dir(here::here(
  "results", "2015-harrison_queen_worker_whole_body", "results",
  "2020-07-14-kallisto"
))

# Read metadata table
table_lifestages_bter <- read.csv(
  here::here(
    "metadata_bter",
    "2015-harrison_queen_worker_whole_body_metadata.csv"
  ),
  header = TRUE,
  stringsAsFactors = TRUE
)

table_lifestages_bter$condition <- relevel(
  table_lifestages_bter$condition,
  ref = "worker_larva"
)

files_lifestages_bter <- here::here(
  "results", "2015-harrison_queen_worker_whole_body",
  "2020-07-14-kallisto", "results",
  table_lifestages_bter$sample,
  "abundance.h5"
)

names(files_lifestages_bter) <- table_lifestages_bter$sample

# Check that sample names match
all(ids_lifestages_bter %in% table_lifestages_bter$sample)

# Check that all files exist
all(file.exists(files_lifestages_bter))
```

## Create a `DESeqDataSet` object
```{r}
# Import kallisto quantifications
txi_lifestages_bter <- tximport::tximport(files_lifestages_bter,
  type = "kallisto",
  tx2gene = tx2gene_bter,
  txOut = FALSE
)

# Create a DESeqDataSet object
dds_lifestages_bter <- DESeq2::DESeqDataSetFromTximport(txi_lifestages_bter,
  colData = table_lifestages_bter,
  design = ~condition
)

# Choose the reference level for condition factor
dds_lifestages_bter$condition <- relevel(dds_lifestages_bter$condition,
  ref = "worker_larva"
)
```

## Pre-filter the low-count genes
```{r}
# Estimate the size factors
dds_lifestages_bter <- DESeq2::estimateSizeFactors(dds_lifestages_bter)

# Pre-filter the dataset
dim(dds_lifestages_bter)
# 10661
# At least 3 samples with a count of 10 or higher
# The number of samples would be set to the smallest group size
keep <- rowSums(DESeq2::counts(dds_lifestages_bter) >= 10) >= 3
dds_lifestages_bter <- dds_lifestages_bter[keep, ]
dim(dds_lifestages_bter)
# 9892
```

## Data transformation and quality assessment
Transform count data using `DESeq2`'s `vst` function. The variance stabilising 
transformation (`vst`) is used for visualization, clustering or other machine 
learning tasks.
```{r}
# Apply the variance stabilising transformation
vsd_lifestages_bter <- DESeq2::vst(dds_lifestages_bter, blind = FALSE)

# Effect of transformation on the variance
vsn::meanSdPlot(assay(vsd_lifestages_bter))

ggsave(
  filename = here(
    "results", "2022-07-20-gene_expression",
    "supplementary", "bter_life_stages_vst_counts_meanSdPlot.pdf"
  ),
  width = 5,
  height = 4,
  units = "in"
)
```

## Plot the (VST) transformed counts
```{r}
# Create a tidy data frame with the transformed counts
vsd_nachrs_lifestages_bter <- assay(vsd_lifestages_bter) %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "ens_gene") %>%
  tidyr::gather(key = "sample", value = "vst_count", -ens_gene) %>%
  dplyr::filter(ens_gene %in% nachrs_bter) %>%
  dplyr::filter(ens_gene != "LOC100645032") %>%
  dplyr::mutate(subunit = case_when(
    ens_gene == "LOC100647624" ~ "alpha1",
    ens_gene == "LOC100647301" ~ "alpha2",
    ens_gene == "LOC100647350" ~ "alpha3",
    ens_gene == "LOC100648987" ~ "alpha4e5",
    ens_gene == "LOC100649515" ~ "alpha5",
    ens_gene == "LOC100643274" ~ "alpha6",
    ens_gene == "LOC100649796" ~ "alpha7",
    ens_gene == "LOC100643282" ~ "alpha8",
    ens_gene == "LOC100649612" ~ "beta1",
    ens_gene == "LOC100646787" ~ "beta2"
  )) %>%
  dplyr::mutate(condition = case_when(
    grepl("queen_adult", sample) ~ "queen_adult",
    grepl("worker_adult", sample) ~ "worker_adult",
    grepl("worker_larva", sample) ~ "worker_larva",
    grepl("worker_pupa", sample) ~ "worker_pupa"
  )) %>%
  dplyr::mutate_if(is.numeric, round, digits = 4)

vsd_nachrs_lifestages_bter %>%
  dplyr::mutate(
    condition = stringr::str_to_sentence(condition)
  ) %>%
  dplyr::mutate(
    condition = stringr::str_replace(condition, "_", " ")
  ) %>%
  ggplot(aes(
    x = condition, y = vst_count, color = condition, fill = condition
  )) +
  geom_boxplot(
    alpha = 0.8, width = 0.4, position = position_dodge(0.8), size = 0.6
  ) +
  geom_jitter(
    shape = 21, fill = "white", size = 1.6, stroke = 0.8,
    position = position_dodge(0.8)
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "plain"),
    plot.subtitle = element_text(hjust = 0.5, size = 14, face = "plain"),
    strip.text = element_text(size = 12, face = "plain", hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(size = 12),
    legend.position = "bottom",
    axis.title = element_text(size = 12),
    axis.text.x = element_blank(),
    panel.spacing.x = unit(1, "mm")
  ) +
  labs(
    x = "Condition",
    y = expression(paste("Log"[2], " (VST) transformed count")) # batch variation removed
  ) +
  facet_wrap(~subunit, ncol = 10) +
  scale_color_brewer(palette = "Set2") +
  scale_fill_brewer(palette = "Set2")

ggsave(
  filename = here::here(
    "results", "2022-07-20-gene_expression",
    "supplementary", "bter_life_stages_vst_counts_nachrs_boxplots.pdf"
  ),
  width = 10,
  height = 5,
  units = "in"
)
```

## Principal component plot of the samples
```{r}
pca_data_lifestages_bter <- DESeq2::plotPCA(vsd_lifestages_bter,
  intgroup = "condition",
  n = 500,
  returnData = TRUE
)

# Customize PCA plot
percent_var_lifestages_bter <- round(
  100 * attr(pca_data_lifestages_bter, "percentVar")
)

pca_data_lifestages_bter %>%
  dplyr::mutate(
    condition = stringr::str_to_sentence(condition)
  ) %>%
  dplyr::mutate(
    condition = stringr::str_replace(condition, "_", " ")
  ) %>%
  ggplot(
    aes(x = PC1, y = PC2, color = condition, shape = condition)
  ) +
  geom_point(size = 4, alpha = 0.8) +
  # geom_point(size = 4, stroke = 1, shape = 21) +
  xlab(paste0("PC1 [", percent_var_lifestages_bter[1], "%]")) +
  ylab(paste0("PC2 [", percent_var_lifestages_bter[2], "%]")) +
  coord_fixed(ratio = 1) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "plain"),
    legend.title = element_blank(),
    legend.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  ) +
  labs(
    title = "Relationships among samples of whole bodies of bumble bees"
  ) +
  scale_color_brewer(palette = "Set2") +
  scale_fill_brewer(palette = "Set2")

ggsave(
  filename = here::here(
    "results", "2022-07-20-gene_expression",
    "supplementary", "bter_life_stages_vst_counts_pca.pdf"
  ),
  width = 8,
  height = 6,
  units = "in"
)
```
