---
title: "Composition analysis of nAChR genes in _Apis mellifera_"
author: "Alicja Witwicka, Federico Lopez"
date: '`r Sys.Date()`'
output:
  github_document:
    toc: yes
  pdf_document: 
    fig_caption: yes
    toc: yes
  html_document:
    toc: yes
number_sections: true

---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache.lazy = FALSE,
  include = FALSE,
  out.height = "\textheight",
  out.width = "\textwidth"
)
```

# Load libraries
```{r}
load_pkgs <- function(pkg) {
  sapply(pkg, require, character.only = TRUE)
}

cran_pkgs <- c(
  "BiocManager", "tidyverse", "GGally", "ggtext", "ggh4x", "ggforce",
  "RColorBrewer", "styler", "patchwork", "here"
)

load_pkgs(cran_pkgs)

bioconductor_pkgs <- c(
  "biomaRt", "rhdf5", "tximport", "DESeq2", "vsn", "hexbin"
)

load_pkgs(bioconductor_pkgs)
```

# Set a custom `ggplot` theme
```{r}
# Set a custom ggplot theme
# devtools::install_github("cttobin/ggthemr")
library(ggthemr)

# Diverging
custom_palette <- c(
  "#9E9E9E", "#88CCEE", "#18913D", "#E9C83A", "#882255", "#CC6677",
  "#332288", "#7B73D1", "#2857BA", "#28A7B6", "#F2CB57", "#F2845C"
)

# ggthemr::colour_plot(custom_palette)

custom_theme <- ggthemr::define_palette(
  swatch = custom_palette,
  gradient = c(lower = "#E9C83A", upper = "#882255")
)

ggthemr::ggthemr(custom_theme)
```

```{r}
# here::here()
# Create directory for results
dir.create(here::here("results", "2022-07-20-gene_expression"),
  recursive = TRUE
)
```

# Read tables of samples and covariates
```{r}
# Jasper et al., 2015
# Set sample identifiers using the names of the kallisto output directories
ids_jasper <- dir(here::here(
  "results", "2015-jasper_worker_tissues", "2020-06-16-kallisto", "results"
))

# Read metadata table
table_jasper <- read.csv(
  here::here(
    "metadata_amel",
    "2015-jasper_worker_tissues_metadata.csv"
  ),
  header = TRUE,
  stringsAsFactors = TRUE
)

files_jasper <- here::here(
  "results", "2015-jasper_worker_tissues", "2020-06-16-kallisto", "results",
  table_jasper$sample,
  "abundance.h5"
)

names(files_jasper) <- table_jasper$sample

# Check that sample names match
all(ids_jasper %in% table_jasper$sample)

# Check that all files exist
all(file.exists(files_jasper))

# Select the worker "brain" samples
jasper_brain_table <- table_jasper %>%
  dplyr::filter(tissue == "brain")

jasper_brain_files <- files_jasper[grep("brain", files_jasper)]

# Manfredini et al., 2015
# Set sample identifiers using the names of the kallisto output directories
ids_manfredini_amel <- dir(here::here(
  "results", "2015-manfredini_queen_brains", "2020-06-18-kallisto", "results"
))

# Read metadata table
table_manfredini_amel <- read.csv(
  here::here(
    "metadata_amel",
    "2015-manfredini_queen_brains_metadata.csv"
  ),
  header = TRUE,
  stringsAsFactors = TRUE
)

files_manfredini_amel <- here::here(
  "results", "2015-manfredini_queen_brains", "2020-06-18-kallisto", "results",
  table_manfredini_amel$sample,
  "abundance.h5"
)

names(files_manfredini_amel) <- table_manfredini_amel$sample

# Check that sample names match
all(ids_manfredini_amel %in% table_manfredini_amel$sample)

# Check that all files exist
all(file.exists(files_manfredini_amel))

# Liberti et al., 2019
# Set sample identifiers using the names of the kallisto output directories
ids_liberti <- dir(here::here(
  "results", "2019-liberti_queen_brains", "2020-06-18-kallisto", "results"
))

# Read metadata table
table_liberti <- read.csv(
  here::here(
    "metadata_amel",
    "2019-liberti_queen_brains_metadata.csv"
  ),
  header = TRUE,
  stringsAsFactors = TRUE
)

files_liberti <- here::here(
  "results", "2019-liberti_queen_brains", "2020-06-18-kallisto", "results",
  table_liberti$sample,
  "abundance.h5"
)

names(files_liberti) <- table_liberti$sample

# Check that sample names match
all(ids_liberti %in% table_liberti$sample)

# Check that all files exist
all(file.exists(files_liberti))

# Christen et al., 2015
# Set sample identifiers using the names of the kallisto output directories
ids_christen <- dir(here::here(
  "results", "2018-christen_worker_brains", "2020-06-18-kallisto", "results"
))

# Read metadata table
table_christen <- read.csv(
  here::here(
    "metadata_amel",
    "2018-christen_worker_brains_metadata.csv"
  ),
  header = TRUE,
  stringsAsFactors = TRUE
)

files_christen <- here::here(
  "results", "2018-christen_worker_brains", "2020-06-18-kallisto", "results",
  table_christen$sample,
  "abundance.h5"
)

names(files_christen) <- table_christen$sample

# Check that sample names match
all(ids_christen %in% table_christen$sample)

# Check that all files exist
all(file.exists(files_christen))
```

# Combine input files and information about the samples
```{r}
# Castes
files_castes_amel <- c(
  jasper_brain_files, files_manfredini_amel, files_liberti, files_christen
)
table_castes_amel <- dplyr::bind_rows(
  jasper_brain_table, table_manfredini_amel, table_liberti, table_christen
) %>%
  dplyr::select(sample, caste, tissue, study, species)

# Tissues
files_tissues_amel <- files_jasper
table_tissues_amel <- table_jasper
```

# Associate transcripts to genes
```{r}
mart_amel <- biomaRt::useMart(
  biomart = "metazoa_mart",
  dataset = "amellifera_eg_gene",
  host = "sep2019-metazoa.ensembl.org" # version 45
)

tx2gene_amel <- biomaRt::getBM(
  attributes = c("ensembl_transcript_id", "ensembl_gene_id"),
  mart = mart_amel
)
```

# Import `kallisto` quantifications using `tximport`
```{r}
# Castes
txi_castes_amel <- tximport::tximport(
  files_castes_amel,
  type = "kallisto",
  tx2gene = tx2gene_amel,
  txOut = FALSE
)

# Tissues
txi_tissues_amel <- tximport::tximport(
  files_tissues_amel,
  type = "kallisto",
  tx2gene = tx2gene_amel,
  txOut = FALSE
)
```

# Create vector of nAChR gene identifiers
```{r}
nachrs_amel <- c(
  "GB42850", "GB42644", "GB47845", "GB43275", "GB43064", "GB43416",
  "GB53053", "GB40923", "GB53427", "GB53055", "GB53428"
)
```

# Analysis of castes
## Create a `DESeqDataSet` object
```{r}
dds_castes_amel <- DESeq2::DESeqDataSetFromTximport(txi_castes_amel,
  colData = table_castes_amel,
  design = ~ caste
)

# Choose the reference level for caste factor
dds_castes_amel$caste <- relevel(dds_castes_amel$caste, ref = "worker")

# Estimate the size factors
dds_castes_amel <- DESeq2::estimateSizeFactors(dds_castes_amel)

# Pre-filter the low-count genes
dim(dds_castes_amel)
# 15314    18
keep <- rowSums(DESeq2::counts(dds_castes_amel) >= 10) >= 3
dds_castes_amel <- dds_castes_amel[keep, ]
dim(dds_castes_amel)
# 12232    18
```

## Use normalised counts to calculate the subunit proportions per sample
```{r}
# Normalise counts
counts_castes_amel <- DESeq2::counts(dds_castes_amel, normalized = TRUE)

counts_nachrs_castes_amel <- counts_castes_amel %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "ens_gene") %>%
  dplyr::filter(ens_gene %in% nachrs_amel) %>%
  tidyr::gather(key = "sample", value = "norm_count", -ens_gene) %>%
  dplyr::group_by(sample) %>%
  dplyr::mutate(proportion = norm_count / sum(norm_count)) %>%
  dplyr::mutate(subunit = case_when(
    ens_gene == "GB42850" ~ "alpha1",
    ens_gene == "GB42644" ~ "alpha2",
    ens_gene == "GB47845" ~ "alpha3",
    ens_gene == "GB43275" ~ "alpha4",
    ens_gene == "GB43064" ~ "alpha5",
    ens_gene == "GB43416" ~ "alpha6",
    ens_gene == "GB53053" ~ "alpha7",
    ens_gene == "GB40923" ~ "alpha8",
    ens_gene == "GB53427" ~ "alpha9",
    ens_gene == "GB53055" ~ "beta1",
    ens_gene == "GB53428" ~ "beta2"
  )) %>%
  dplyr::mutate(caste = case_when(
    grepl("n\\d|f\\d", sample) ~ "worker",
    grepl("worker", sample) ~ "worker",
    grepl("queen", sample) ~ "queen"
  )) %>%
  dplyr::mutate_if(is.numeric, round, digits = 4) %>%
  dplyr::ungroup()
```

## Fit glm with subunit proportion as the response variable
```{r}
# Fit glm
glmfit_castes_amel <- glm(
  formula = proportion ~ ens_gene + caste + ens_gene * caste,
  family = quasibinomial(link = "logit"), data = counts_nachrs_castes_amel
)

# Save summary
sink(here::here(
  "results", "2022-07-20-gene_expression",
  "amel_castes_proportions_glm_summary.txt"
))
print(summary(glmfit_castes_amel))
sink() # returns output to the console

# Save glm results as a tidy tibble
glmfit_castes_tidy_amel <- broom::tidy(glmfit_castes_amel) %>%
  dplyr::mutate_if(is.numeric, round, digits = 4)

write.csv(glmfit_castes_tidy_amel,
  file = here::here(
    "results", "2022-07-20-gene_expression",
    "amel_castes_proportions_glm_tidy.csv"
  ),
  row.names = FALSE
)
```

# Analysis of tissues
## Create a `DESeqDataSet` object
```{r}
dds_tissues_amel <- DESeq2::DESeqDataSetFromTximport(txi_tissues_amel,
  colData = table_tissues_amel,
  design = ~ tissue
)

# Choose the reference level for tissue factor
dds_tissues_amel$tissue <- relevel(dds_tissues_amel$tissue, ref = "midgut")
```

## Pre-filter the low-count genes
```{r}
# Estimate the size factors
dds_tissues_amel <- DESeq2::estimateSizeFactors(dds_tissues_amel)

# Pre-filter the dataset
dim(dds_tissues_amel)
# 15314    48
keep <- rowSums(DESeq2::counts(dds_tissues_amel) >= 10) >= 3
dds_tissues_amel <- dds_tissues_amel[keep, ]
dim(dds_tissues_amel)
# 13114    48
```

## Use normalised counts to calculate the subunit proportions per sample
```{r}
# Normalise counts
counts_tissues_amel <- DESeq2::counts(dds_tissues_amel, normalized = TRUE)

counts_nachrs_tissues_amel <- counts_tissues_amel %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "ens_gene") %>%
  dplyr::filter(ens_gene %in% nachrs_amel) %>%
  tidyr::gather(key = "sample", value = "norm_count", -ens_gene) %>%
  dplyr::group_by(sample) %>%
  dplyr::mutate(proportion = norm_count / sum(norm_count)) %>%
  dplyr::mutate(subunit = case_when(
    ens_gene == "GB42850" ~ "alpha1",
    ens_gene == "GB42644" ~ "alpha2",
    ens_gene == "GB47845" ~ "alpha3",
    ens_gene == "GB43275" ~ "alpha4",
    ens_gene == "GB43064" ~ "alpha5",
    ens_gene == "GB43416" ~ "alpha6",
    ens_gene == "GB53053" ~ "alpha7",
    ens_gene == "GB40923" ~ "alpha8",
    ens_gene == "GB53427" ~ "alpha9",
    ens_gene == "GB53055" ~ "beta1",
    ens_gene == "GB53428" ~ "beta2"
  )) %>%
  dplyr::mutate(tissue = case_when(
    grepl("brain", sample) ~ "brain",
    grepl("antenna", sample) ~ "antenna",
    grepl("midgut", sample) ~ "midgut",
    grepl("hypopharyngeal_gland", sample) ~ "hypopharyngeal_gland",
    grepl("malpighian_tubule", sample) ~ "malpighian_tubule",
    grepl("mandibular_gland", sample) ~ "mandibular_gland",
    grepl("muscle", sample) ~ "muscle",
    grepl("nasonov_gland", sample) ~ "nasonov_gland",
    grepl("thoracic_ganglion", sample) ~ "thoracic_ganglion"
  )) %>%
  dplyr::mutate(caste = "worker") %>%
  dplyr::mutate_if(is.numeric, round, digits = 4)
```

## Fit glm with subunit proportion as the response variable
```{r}
# Fit glm
glmfit_tissues_amel <- glm(
  formula = proportion ~ ens_gene + tissue + ens_gene * tissue,
  family = quasibinomial(link = "logit"), data = counts_nachrs_tissues_amel
)

# Save summary
sink(here::here(
  "results", "2022-07-20-gene_expression",
  "amel_tissues_proportions_glm_summary.txt"
))
print(summary(glmfit_tissues_amel))
sink() # returns output to the console

# Save glm results as a tidy tibble
glmfit_tissues_tidy_amel <- broom::tidy(glmfit_tissues_amel) %>%
  dplyr::mutate_if(is.numeric, round, digits = 4)

write.csv(glmfit_tissues_tidy_amel,
  file = here::here(
    "results", "2022-07-20-gene_expression",
    "amel_tissues_proportions_glm_tidy.csv"
  ),
  row.names = FALSE
)
```

# Supplementary materials: data visualisation
## Stacked bar chart of subunit proportions in tissues
```{r}
levels_tissues_amel <- c(
  "brain", "thoracic_ganglion", "antenna",
  "hypopharyngeal_gland", "mandibular_gland", "muscle",
  "midgut", "malpighian_tubule", "nasonov_gland"
)

levels_subunits <- c(
  "alpha1", "alpha2", "alpha3", "alpha4", "alpha5", "alpha6",
  "alpha7", "alpha8", "alpha9", "beta1", "beta2"
)

counts_nachrs_tissues_amel %>%
  dplyr::mutate(
    tissue = stringr::str_to_sentence(tissue)
  ) %>%
  dplyr::mutate(
    tissue = stringr::str_replace(tissue, "_", " ")
  ) %>%
  ggplot(aes(
    x = factor(tissue,
      levels = stringr::str_to_sentence(levels_tissues_amel) %>%
        stringr::str_replace("_", " ")
    ),
    y = proportion,
    fill = factor(subunit, levels = levels_subunits),
    color = factor(subunit, levels = levels_subunits)
  )) +
  geom_bar(position = "fill", stat = "identity", width = 0.5, alpha = 0.9) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "plain"),
    plot.subtitle = element_text(hjust = 0.5, size = 12, face = "plain"),
    axis.title = element_text(size = 14),
    legend.title = element_blank(),
    strip.text = element_blank(),
    axis.text = element_text(size = 12),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    legend.text = element_text(size = 12),
    legend.key.height = unit(0.5, "cm"),
    legend.key.width = unit(0.5, "cm")
  ) +
  labs(
    x = element_blank(),
    y = "Proportion of total nAChR expression"
  )
```
