---
title: "Composition analysis of nAChR genes in _Apis mellifera_ using `DESeq2`"
author: "Alicja Witwicka"
date: '`r Sys.Date()`'
output:
  github_document:
    toc: yes
  pdf_document: 
    fig_caption: yes
    toc: yes
  html_document:
    toc: yes
number_sections: true

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.path = "")
```
```{r libraries, echo = FALSE, message = FALSE, warning = FALSE}
# renv::snapshot()
```
```{r libraries, echo = FALSE, message = FALSE, warning = FALSE}
# Load all packages
library("tidyverse")
library("tximport")
library("readr")
library("ggplot2")
library("DESeq2")
library("dplyr")
library("tibble")
library("limma")
library("ggthemr")
for (package in (.packages())) {
  print(paste("Package", package, "version", packageVersion(package)))
}
```
```{r colours, include = FALSE}
# Set colours
custom_palette <- c(
  "#9E9E9E", "#59BDEF", "#EBCC2A", "#E1AF00", "#F27C8D", "#7B73D1", "#7B9FE0",
  "#F5CC7F", "#66C2A5", "#28A7B6", "#F2CB57", "#F2A057", "#F2845C"
)

# ggthemr::colour_plot(custom_palette)

ggthemr::colour_plot(custom_palette)
custom_theme <- ggthemr::define_palette(
  swatch = custom_palette,
  gradient = c(lower = "#59BDEF", upper = "#FF6B5A")
)
ggthemr::ggthemr(custom_theme)
```

## Introduction

# <b>Questions:</b>
  1. What is the ratio of subunits used in different tissues and castes in Apis mellifera.

Datasets used:
<br><b>worker brains & other tissues</b>
<br>-Jasper et al. 2015
<br>-Christen et al. 2018
<br><b>queen brains</b>
<br>-Liberti et al. 2015
<br>-Manfredini et al. 2015

## Load data and prepare it for analysis
- The data was loaded using tximport, metadata (.csv) was composed and imported

```{r import_files_&_metadata, include = FALSE}
### Read kallisto output and metadata ###
# Set sample identifiers using the names of the kallisto output directories
liberti_ids <- dir(file.path("results/2015-liberti_queen_brains/2020-06-18-kallisto/results"))
manfredini_ids <- dir(file.path("results/2015-manfredini_queen_brains/2020-06-18-kallisto/results"))
christen_ids <- dir(file.path("results/2018-christen_worker_brains/2020-06-18-kallisto/results"))
jasper_ids <- dir(file.path("results/2015-jasper_worker_tissues/2020-06-16-kallisto/results/"))
jasper_ids_brains <- tail(jasper_ids, 6) # select brains for the analysis of castes

# Create sample ids vectors
castes_ids <- c(liberti_ids, christen_ids, manfredini_ids, jasper_ids_brains)
tissues_ids <- jasper_ids

# Read metadata table
## castes
samples_table_castes <- read.csv(
  file.path("amel_metadata/2020_amel_castes_brain.csv"),
  header = TRUE,
  stringsAsFactors = TRUE
)
## tissues
samples_table_tissues <- read.csv(file.path("amel_metadata/2015-jasper_worker_tissues_metadata.csv"),
  header = TRUE,
  stringsAsFactors = TRUE
)

# Read .h5 files in
## castes
liberti_filenames <- file.path(
  "results/2015-liberti_queen_brains/2020-06-18-kallisto/results",
  liberti_ids, "abundance.h5"
)
manfredini_filenames <- file.path(
  "results/2015-manfredini_queen_brains/2020-06-18-kallisto/results",
  manfredini_ids, "abundance.h5"
)
christen_filenames <- file.path(
  "results/2018-christen_worker_brains/2020-06-18-kallisto/results",
  christen_ids, "abundance.h5"
)
jasper_filenames <- file.path(
  "results/2015-jasper_worker_tissues/2020-06-16-kallisto/results/",
  jasper_ids_brains, "abundance.h5"
)
castes_filenames <- c(liberti_filenames, christen_filenames, manfredini_filenames, jasper_filenames)

## tissues
tissues_filenames <- file.path(
  "results/2015-jasper_worker_tissues/2020-06-16-kallisto/results/",
  tissues_ids, "abundance.h5"
)

# Assign names to files
## castes
names(castes_filenames) <- samples_table_castes$sample
## tissues
names(tissues_filenames) <- samples_table_tissues$sample

# check that all files were assigned
all(file.exists(castes_filenames))
all(file.exists(tissues_filenames))
```
- I used biomaRt to associate transcripts to genes (from ensmbl metazoa)
- I imported Kallisto output
```{r biomaRt, include = FALSE}
#### Associate transcripts to genes ####
biomaRt::listMarts(host = "metazoa.ensembl.org")
# biomaRt::listMarts(host = "eg40-metazoa.ensembl.org") # version 40
# biomaRt::listEnsemblArchives()
metazoa_mart <- biomaRt::useMart(
  biomart = "metazoa_mart",
  dataset = "amellifera_eg_gene",
  host = "metazoa.ensembl.org"
)

transcript_to_gene <- biomaRt::getBM(
  attributes = c("ensembl_transcript_id", "ensembl_gene_id"),
  mart = metazoa_mart
)

# Remove all biomaRt cached files
# biomaRt::biomartCacheInfo()
# biomaRt::biomartCacheClear()
# biomaRt::listAttributes(metazoa_mart)
```
```{r tximport, include = FALSE}
# Import kallisto quantifications
## castes
txi_kallisto_castes <- tximport::tximport(
  castes_filenames,
  type = "kallisto",
  tx2gene = transcript_to_gene,
  txOut = FALSE
)
## tissues
txi_kallisto_tissues <- tximport::tximport(
  tissues_filenames,
  type = "kallisto",
  tx2gene = transcript_to_gene,
  txOut = FALSE
)
```
# Create vector of nAChR gene identifiers
```{r gene_vector, message = FALSE}
gene_id <- c(
  "GB40923", "GB53053", "GB42644", "GB53055", "GB47845", "GB43416",
  "GB42850", "GB43275", "GB53427", "GB53428", "GB50159"
)
names_AChRs <- c(
  "alpha8", "alpha7", "alpha2", "beta1", "alpha3", "alpha6", "alpha1",
  "alpha4", "alpha9", "beta2", "alpha5"
)
```

### Analisis of castes 
# Studies:
Datasets used:
<br><b>worker brains</b>
<br>-Jasper et al. 2015
<br>-Christen et al. 2018
<br><b>queen brains</b>
<br>-Liberti et al. 2015
<br>-Manfredini et al. 2015

- Before using DESeq2 I remove unvanted study using RUVSeq

```{r RUVSeq, message= FALSE}

# Set variables for the model
caste <- as.character(unique(samples_table_castes$sample))
study <- as.character(unique(samples_table_tissues$sample))

# Create a DESeq object design: ~caste
dds_castes <- DESeq2::DESeqDataSetFromTximport(txi_kallisto_castes,
  colData = samples_table_castes,
  design = ~caste
)
# Select expressed genes
dim(dds_castes) # 15314 18
# At least 4 samples with a count of 10 or higher
# The number of samples would be set to the smallest group size
keep <- rowSums(DESeq2::counts(dds_castes) >= 10) >= 4
dds_castes <- dds_castes[keep, ]
dim(dds_castes) # 12042 18

# Run DESeq2 pipeline prior to RUVSeq
castes_dds1 <- DESeq2::DESeq(dds_castes, test = "Wald", betaPrior = FALSE)
DESeq2::resultsNames(castes_dds1)
castes_results1 <- DESeq2::results(castes_dds1)

# Run RUVSeq (remove unwanted variance)
## Extract raw counts from dds object
castes_set <- EDASeq::newSeqExpressionSet(DESeq2::counts(dds_castes))
## Normalize the data using upper-quartile (UQ) normalization
castes_set <- EDASeq::betweenLaneNormalization(castes_set, which = "upper")
## Select a set of genes which are not signifficant - not affected by the batch (by defaoul can be homekeeping genes)
not.sig <- rownames(castes_results1)[which(castes_results1$pvalue > 0.4)]
empirical <- rownames(castes_set)[rownames(castes_set) %in% not.sig]
## RUVg uses negative control genes, assumed to have constant expression across samples
castes_set <- RUVSeq::RUVg(castes_set, empirical, k = 1)
# Calculate covariant for the new model (accounting for the study effect)
covariant <- pData(castes_set)

# Transform for ggplot2
cov_1 <- covariant[, 1] %>%
  as.data.frame()
rownames(cov_1) <- samples_table_castes[, 1]
study <- data.frame(samples_table_castes[, 4])
rownames(study) <- samples_table_castes[, 1]
df_merged <- merge(cov_1, study, by = 0)

# Boxplot
ggplot(df_merged, aes(x = samples_table_castes...4., y = .)) +
  geom_boxplot(aes(colour = samples_table_castes...4.), fill = "white") +
  geom_jitter(aes(colour = samples_table_castes...4.))

# Test if extrapolated variables overlap with the study effect
summary(glm(df_merged$. ~ df_merged$samples_table_castes...4.))
```

```{r DESeq2_results, message=FALSE}

# Add calculated covariate to metadata table
colnames(cov_1) <- c("covariate")
cov_1$sample <- samples_table_castes[, 1]
merged_samples_table <- merge(samples_table_castes, cov_1, by = "sample")


## RUN DESEQ2 AGAIN
cast <- as.character(unique(merged_samples_table$condition))
study <- as.character(unique(merged_samples_table$study))
covariate <- as.numeric(unique(merged_samples_table$covariate))

# Create a DESeq object
dds_castes2 <- DESeq2::DESeqDataSetFromTximport(txi_kallisto_castes,
  colData = merged_samples_table,
  design = ~ caste + covariate
)
```
Use normalised counts to calculate the expression proportions per sample.
```{r}

dds_castes2 <- DESeq2::estimateSizeFactors(dds_castes2)
dds_castes2 <- DESeq2::counts(dds_castes2, normalized = TRUE)
# Pre-filter the dataset
dim(dds_castes2)
keep <- rowSums(dds_castes2 >= 10) >= 4
dds_castes2 <- dds_castes2[keep, ]
dim(dds_castes2)

# Extract genes of interest
gene_list_castes <- dds_castes2[gene_id, ]
row.names(gene_list_castes) <- names_AChRs
```
# Normalising TPMs

```{r normalisation, message = FALSE}
# Calculate sums of each collumn
col_sum_castes <- as.data.frame(gene_list_castes) %>% dplyr::summarize_if(is.numeric, sum)

ratios_castes <- mapply("/", as.data.frame(gene_list_castes), col_sum_castes)
row.names(ratios_castes) <- names_AChRs
ratios_castes
```


# Plotting the data

```{r visualisation, message = FALSE}
# Transform data frame
ratios_castes <- as.data.frame(ratios_castes)
ratios_transf_castes <- ratios_castes[, order(names(ratios_castes))]
plot.list_castes <- ratios_transf_castes %>% tibble::rownames_to_column("AChRs")
plot.list_castes <- data.frame(plot.list_castes[1], stack(plot.list_castes[2:ncol(plot.list_castes)]))
col_names_plot <- c("nAChRs", "tpm", "castes")
colnames(plot.list_castes) <- col_names_plot
# create a df of samples which I lated add up to the plot.list
sample <- plot.list_castes$castes
# Remove repeating pattern in castes
plot.list_castes <- plot.list_castes %>%
  mutate_all(~ gsub("abundance.|_1|_2|_3|_4|_5|_n1|_n2|_n3|_f1|_f2|_f3", "", .)) %>%
  mutate_all(~ gsub("_virgin|_sugar_exposed|_mock_insemination", "", .))
plot.list_castes$castes <- gsub("SRR.{6,7}_", "", plot.list_castes$castes)
plot.list_castes$castes <- gsub("brain", "worker", plot.list_castes$castes)
# Make TPMs numeric
plot.list_castes$tpm <- as.numeric(plot.list_castes$tpm)
# Add sample
plot.list_castes$sample <- sample

# Plot data
ggplot(plot.list_castes, aes(x = castes, y = tpm, fill = nAChRs)) +
  geom_bar(width = 0.4, stat = "identity", position = "fill") +
  xlab("castes") +
  ylab("Proportion of all nAChRs expression") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "plain"),
    plot.subtitle = element_text(hjust = 0.5, size = 12, face = "plain"),
    strip.text = element_text(size = 10, face = "plain", hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(size = 10, angle = 65, hjust = 1),
    aspect.ratio = 1 / 1.5
  )
```
```{r logit, message = FALSE}

# Trying to replace alpha8 with aalpha8 (a8 as a reference, logit takes them alphabetically, didn't know how to do this otherwise)
plot.list_castes <- plot.list_castes %>% mutate(nAChRs = ifelse(as.character(nAChRs) == "aalpha8", "alpha8", as.character(nAChRs)))

# Run model
logistic <- glm(tpm ~ nAChRs * castes, data = plot.list_castes, family = quasibinomial(link = "logit"))
summary(logistic)

# Turn glm results into a tidy tibble
broom::tidy(logistic)
# Look at the diagnostics
broom::glance(logistic)
# Return a tibble with information about data points
broom::augment(logistic)
```

### Analisis of tissues
# Studies:
Datasets used:
<br><b>9 worker (nurses and foragers) tissues</b>
<br>-Jasper et al. 2015

```{r filtering, message = FALSE, warning = FALSE}

# Set variables for the model (tissues)
condition <- samples_table_tissues %>%
  dplyr::pull(tissue) %>%
  unique() %>%
  as.character()

# Set the reference level: mitgut
samples_table_tissues$condition <- relevel(
  samples_table_tissues$tissue,
  ref = "midgut"
)

# Create a DESeq object (for mitgut comparison)
dds_tissues <- DESeq2::DESeqDataSetFromTximport(txi_kallisto_tissues,
  colData = samples_table_tissues,
  design = ~tissue
)
dds_tissues <- DESeq2::estimateSizeFactors(dds_tissues)
dds_tissues <- DESeq2::counts(dds_tissues, normalized = TRUE)
# Pre-filter the dataset
dim(dds_tissues) # 15314    48
keep <- rowSums(dds_tissues >= 10) >= 4
dds_tissues <- dds_tissues[keep, ]
dim(dds_tissues) # 1 2899    48


# Extract genes of interest
gene_list_tissues <- dds_tissues[gene_id, ]
row.names(gene_list_tissues) <- names_AChRs
```
# Normalising TPMs

```{r normalisation, message = FALSE}
# Calculate sums of each collumn
col_sum_tissues <- as.data.frame(gene_list_tissues) %>% dplyr::summarize_if(is.numeric, sum)

ratios_tissues <- mapply("/", as.data.frame(gene_list_tissues), col_sum_tissues)
row.names(ratios_tissues) <- names_AChRs
```


# Plotting the data

```{r ratios visualisation, message = FALSE}
# Transform data frame
ratios_tissues <- as.data.frame(ratios_tissues)
ratios_transf_tissues <- ratios_tissues[, order(names(ratios_tissues))]
plot.list_tissues <- ratios_transf_tissues %>% tibble::rownames_to_column("AChRs")
plot.list_tissues <- data.frame(plot.list_tissues[1], stack(plot.list_tissues[2:ncol(plot.list_tissues)]))
colnames(plot.list_tissues) <- c("nAChRs", "tpm", "tissue")

# Remove repeating pattern in tissue names
plot.list_tissues <- plot.list_tissues %>% mutate_all(~ gsub("abundance.|_f1|_f2|_f3|_n1|_n2|_n3", "", .))
plot.list_tissues$tissue <- gsub("SRR.{6,7}_", "", plot.list_tissues$tissue)
plot.list_tissues$tissue <- gsub("thoracic_ganglia", "thoracic_ganglion", plot.list_tissues$tissue)
# Make TPMs numeric
plot.list_tissues$tpm <- as.numeric(plot.list_tissues$tpm)

# Plot data
ggplot(plot.list_tissues, aes(x = tissue, y = tpm, fill = nAChRs)) +
  geom_bar(width = 0.7, stat = "identity", position = "fill") +
  xlab("Tissue") +
  ylab("Proportion of total nAChRs expression") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "plain"),
    plot.subtitle = element_text(hjust = 0.5, size = 12, face = "plain"),
    strip.text = element_text(size = 10, face = "plain", hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(size = 10, angle = 65, hjust = 1),
    aspect.ratio = 1 / 1.5
  )
```
```{r logit, message = FALSE}


# Rearrange references
replace_brain <- plot.list_tissues %>% mutate(tissue = ifelse(as.character(tissue) == "brain", "aabrain", as.character(tissue)))
replace_a8 <- replace_brain %>% mutate(nAChRs = ifelse(as.character(nAChRs) == "alpha8", "aalpha8", as.character(nAChRs)))

logistic <- glm(tpm ~ nAChRs * tissue, data = replace_a8, family = quasibinomial(link = "logit"))
summary(logistic)
# Turn glm results into a tidy tibble
broom::tidy(logistic)
# Look at the diagnostics
broom::glance(logistic)
# Return a tibble with information about data points
broom::augment(logistic)
```
