---
title: "Correlation tests between _A. mellifera_ and _B. terrestris_ "
author: "Federico Lopez, Alicja Witwicka"
date: '`r Sys.Date()`'
output:
  github_document:
    toc: yes
  pdf_document: 
    fig_caption: yes
    toc: yes
  html_document:
    toc: yes
editor_options: 
  chunk_output_type: console
geometry: margin = 1cm
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache.lazy = FALSE,
  include = FALSE,
  out.height = "\textheight",
  out.width = "\textwidth"
)
```

# Load libraries
```{r}
load_pkgs <- function(pkg) {
  sapply(pkg, require, character.only = TRUE)
}

cran_pkgs <- c(
  "BiocManager", "tidyverse", "GGally", "RColorBrewer", "pheatmap", "styler",
  "here"
)
load_pkgs(cran_pkgs)

bioconductor_pkgs <- c(
  "biomaRt", "rhdf5", "tximport", "DESeq2", "vsn", "hexbin"
)
load_pkgs(bioconductor_pkgs)
```

# Set a custom `ggplot` theme
```{r}
# Set a custom ggplot theme
library(ggthemr)
custom_palette <- c(
  "#9E9E9E", "#59BDEF", "#EBCC2A", "#E1AF00", "#F27C8D", "#7B73D1", "#7B9FE0",
  "#F5CC7F", "#66C2A5", "#28A7B6", "#F2CB57", "#F2A057", "#F2845C"
)
# ggthemr::colour_plot(custom_palette)

custom_theme <- ggthemr::define_palette(
  swatch = custom_palette,
  gradient = c(lower = "#59BDEF", upper = "#FF6B5A")
)
ggthemr::ggthemr(custom_theme)
```

Create `DESeqDataSet` objects with the samples from brains of _Apis mellifera_
and _B terrestris_.  

Worker caste data sets:  
Jasper et al. 2015 (amel)  
Porath et al. 2019 (bter)  

Queen caste data sets:   
Manfredini et al. 2015 (amel)  
Manfredini et al. 2017 (bter)  

# Read tables of samples and covariates
## _Apis mellifera_
```{r}
# Jasper et al. 2015
# Set sample identifiers using the names of the kallisto output directories
jasper_ids <- dir(here::here(
  "2015-jasper_worker_tissues", "2020-06-16-kallisto", "results"
))

# Read metadata table
jasper_table <- read.csv(
  here::here(
    "amel_metadata",
    "2015-jasper_worker_tissues_metadata.csv"
  ),
  header = TRUE,
  stringsAsFactors = TRUE
)

jasper_files <- here::here(
  "2015-jasper_worker_tissues", "2020-06-16-kallisto", "results",
  jasper_table$sample,
  "abundance.h5"
)

names(jasper_files) <- jasper_table$sample

# Check that sample names match
all(jasper_ids %in% jasper_table$sample)

# Check that all files exist
all(file.exists(jasper_files))

# Select the worker "brain" samples
jasper_brain_table <- jasper_table %>%
  dplyr::filter(tissue == "brain")

jasper_brain_files <- jasper_files[grep("brain", jasper_files)]

# Manfredini et al. 2015
# Set sample identifiers using the names of the kallisto output directories
amel_manfredini_ids <- dir(here::here(
  "2015-manfredini_queen_brains", "2020-06-18-kallisto", "results"
))

# Read metadata table
amel_manfredini_table <- read.csv(
  here::here(
    "amel_metadata",
    "2015-manfredini_queen_brains_metadata.csv"
  ),
  header = TRUE,
  stringsAsFactors = TRUE
)

amel_manfredini_files <- here::here(
  "2015-manfredini_queen_brains", "2020-06-18-kallisto", "results",
  amel_manfredini_table$sample,
  "abundance.h5"
)

names(amel_manfredini_files) <- amel_manfredini_table$sample

# Check that sample names match
all(amel_manfredini_ids %in% amel_manfredini_table$sample)

# Check that all files exist
all(file.exists(amel_manfredini_files))

# Combine input files and information about the samples
amel_brain_files <- c(jasper_brain_files, amel_manfredini_files)
amel_brain_table <- dplyr::bind_rows(
  jasper_brain_table, amel_manfredini_table
) %>%
  dplyr::select(sample, caste, tissue, study, species)
```

## _Bombus terrestris_
```{r}
# Workers
# Porath et al. 2019
# Set sample identifiers using the names of the kallisto output directories
porath_ids <- dir(here::here(
  "2019-porath_worker_brain", "2020-07-14-kallisto", "results"
))

# Read metadata table
porath_table <- read.csv(
  here::here(
    "bter_metadata",
    "2019-porath_worker_brains_metadata.csv"
  ),
  header = TRUE,
  stringsAsFactors = TRUE
)

porath_files <- here::here(
  "2019-porath_worker_brain", "2020-07-14-kallisto", "results",
  porath_table$sample,
  "abundance.h5"
)

names(porath_files) <- porath_table$sample

# Check that sample names match
all(porath_ids %in% porath_table$sample)

# Check that all files exist
all(file.exists(porath_files))

# Queens
# Manfredini et al. 2017
# Set sample identifiers using the names of the kallisto output directories
bter_manfredini_ids <- dir(here::here(
  "2017-manfredini_queen_brain", "2020-08-08-kallisto", "results"
))

# Read metadata table
bter_manfredini_table <- read.csv(
  here::here("bter_metadata", "2017-manfredini_queen_brain_metadata.csv"),
  header = TRUE,
  stringsAsFactors = TRUE
)

# Read .h5 files in
bter_manfredini_files <- here::here(
  "2017-manfredini_queen_brain", "2020-08-08-kallisto", "results",
  bter_manfredini_table$sample,
  "abundance.h5"
)

# Assign names to files
names(bter_manfredini_files) <- bter_manfredini_table$sample

# Check that sample names match
all(bter_manfredini_ids %in% bter_manfredini_table$sample)

# Check that all files exist
all(file.exists(bter_manfredini_files))

# Combine input files and information about the samples
bter_brain_files <- c(porath_files, bter_manfredini_files)
bter_brain_table <- dplyr::bind_rows(
  porath_table, bter_manfredini_table
)
```

# Associate transcripts to genes
```{r}
# Remove all biomaRt cached files
# biomaRt::biomartCacheInfo()
# biomaRt::biomartCacheClear()

# Retrieve Ensembl gene identifiers/names and create transcript-to-gene table
biomaRt::listMarts(host = "metazoa.ensembl.org")
# biomaRt::listMarts(host = "eg40-metazoa.ensembl.org") # version 40
# biomaRt::listEnsemblArchives()

# Apis mellifera
amel_mart <- biomaRt::useMart(
  biomart = "metazoa_mart",
  dataset = "amellifera_eg_gene",
  host = "metazoa.ensembl.org" # version 47
)

amel_tx2gene <- biomaRt::getBM(
  attributes = c("ensembl_transcript_id", "ensembl_gene_id"),
  mart = amel_mart
)

# Bombus terrestris
bter_mart <- biomaRt::useMart(
  biomart = "metazoa_mart",
  dataset = "bterrestris_eg_gene",
  host = "metazoa.ensembl.org" # version 47
)

bter_tx2gene <- biomaRt::getBM(
  attributes = c("ensembl_transcript_id", "ensembl_gene_id"),
  mart = bter_mart
)
```

# Import `kallisto` quantifications using `tximport`
```{r}
# Apis mellifera
amel_brain_txi <- tximport::tximport(amel_brain_files,
  type = "kallisto",
  tx2gene = amel_tx2gene,
  txOut = FALSE
)

# Bombus terrestris
bter_brain_txi <- tximport::tximport(bter_brain_files,
  type = "kallisto",
  tx2gene = bter_tx2gene,
  txOut = FALSE
)
```

# Create `DESeqDataSet` objects
```{r}
# Apis mellifera
amel_brain_dds <- DESeq2::DESeqDataSetFromTximport(amel_brain_txi,
  colData = amel_brain_table,
  design = ~caste
)

# Estimate the size factors
amel_brain_dds <- DESeq2::estimateSizeFactors(amel_brain_dds)
# Filtering
keep <- rowSums(DESeq2::counts(amel_brain_dds) >= 10) >= 3
dds <- amel_brain_dds[keep, ]

# Bombus terrestris
bter_brain_dds <- DESeq2::DESeqDataSetFromTximport(bter_brain_txi,
  colData = bter_brain_table,
  design = ~caste
)

# Estimate the size factors
bter_brain_dds <- DESeq2::estimateSizeFactors(bter_brain_dds)
# Filtering
keep <- rowSums(DESeq2::counts(bter_brain_dds) >= 10) >= 3
dds <- bter_brain_dds[keep, ]
```

# Normalise the counts using `DESeq2`'s median-of-ratios method
```{r}
amel_brain_counts <- DESeq2::counts(amel_brain_dds, normalized = TRUE)
bter_brain_counts <- DESeq2::counts(bter_brain_dds, normalized = TRUE)
```

# Create vectors of nAChR gene identifiers
```{r}
amel_nachrs <- c(
  "GB42850", "GB42644", "GB47845", "GB43275", "GB43064", "GB43416", 
  "GB53053", "GB40923", "GB53427", "GB53055", "GB53428"
)

bter_nachrs <- c(
  "LOC100643274", "LOC100643282", "LOC100645032", "LOC100646787",
  "LOC100647301", "LOC100647350", "LOC100647624", "LOC100648987",
  "LOC100649515", "LOC100649612", "LOC100649796"
)
```

# Create data frames of normalised counts 
## Apis mellifera
```{r}
amel_brain_counts_nachrs <- amel_brain_counts %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "ens_gene") %>%
  tidyr::gather(key = "sample", value = "norm_count", -ens_gene) %>%
  dplyr::mutate_if(is.numeric, round, digits = 4) %>%
  dplyr::filter(ens_gene %in% amel_nachrs) %>%
  dplyr::mutate(subunit = case_when(
    ens_gene == "GB42850" ~ "alpha1",
    ens_gene == "GB42644" ~ "alpha2",
    ens_gene == "GB47845" ~ "alpha3",
    ens_gene == "GB43275" ~ "alpha4",
    ens_gene == "GB43064" ~ "alpha5",
    ens_gene == "GB43416" ~ "alpha6",
    ens_gene == "GB53053" ~ "alpha7",
    ens_gene == "GB40923" ~ "alpha8",
    ens_gene == "GB53427" ~ "alpha9",
    ens_gene == "GB53055" ~ "beta1",
    ens_gene == "GB53428" ~ "beta2"
  )) %>%
  dplyr::mutate(study = case_when(
    stringr::str_detect(
      sample, paste(jasper_table$sample, collapse = "|")
    ) ~ "jasper",
    stringr::str_detect(
      sample, paste(amel_manfredini_table$sample, collapse = "|")
    ) ~ "manfredini"
  )) %>%
  dplyr::mutate(caste = case_when(
    stringr::str_detect(
      sample, paste(jasper_table$sample, collapse = "|")
    ) ~ "worker",
    stringr::str_detect(
      sample, paste(amel_manfredini_table$sample, collapse = "|")
    ) ~ "queen"
  )) %>%
  dplyr::mutate(tissue = "brain") %>%
  dplyr::mutate(species = "amel")
```

## _Bombus terrestris_
```{r}
# "LOC100645032" has low expression
bter_brain_counts_nachrs <- bter_brain_counts %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "ens_gene") %>%
  tidyr::gather(key = "sample", value = "norm_count", -ens_gene) %>%
  dplyr::mutate_if(is.numeric, round, digits = 4) %>%
  dplyr::filter(ens_gene %in% bter_nachrs) %>%
  dplyr::filter(ens_gene != "LOC100645032") %>%
  dplyr::mutate(subunit = case_when(
    ens_gene == "LOC100647624" ~ "alpha1",
    ens_gene == "LOC100647301" ~ "alpha2",
    ens_gene == "LOC100647350" ~ "alpha3",
    ens_gene == "LOC100648987" ~ "alpha4e5",
    ens_gene == "LOC100649515" ~ "alpha5",
    ens_gene == "LOC100643274" ~ "alpha6",
    ens_gene == "LOC100649796" ~ "alpha7",
    ens_gene == "LOC100643282" ~ "alpha8",
    ens_gene == "LOC100649612" ~ "beta1",
    ens_gene == "LOC100646787" ~ "beta2"
  )) %>%
  dplyr::mutate(study = case_when(
    stringr::str_detect(
      sample, paste(porath_table$sample, collapse = "|")
    ) ~ "porath",
    stringr::str_detect(
      sample, paste(bter_manfredini_table$sample, collapse = "|")
    ) ~ "manfredini"
  )) %>%
  dplyr::mutate(caste = case_when(
    stringr::str_detect(
      sample, paste(porath_table$sample, collapse = "|")
    ) ~ "worker",
    stringr::str_detect(
      sample, paste(bter_manfredini_table$sample, collapse = "|")
    ) ~ "queen"
  )) %>%
  dplyr::mutate(tissue = "brain") %>%
  dplyr::mutate(species = "bter")
```

# Combine data frames of normalised counts
```{r}
amel_bter_brain_counts_nachrs <- dplyr::bind_rows(
  amel_brain_counts_nachrs, bter_brain_counts_nachrs
)

# Recode subunit "alpha4" in Bombus terrestris
amel_bter_brain_counts_nachrs$subunit <- dplyr::recode(
  amel_bter_brain_counts_nachrs$subunit,
  "alpha4e5" = "alpha4"
)
```

# Calculate the medians of subunit proportions
```{r}
# Create data frames of normalised counts and calculate expression proportions
# Discard data from divergent subunits
divergent_subunits <- c("alpha9", "beta2")

amel_bter_brain_proportions <- amel_bter_brain_counts_nachrs %>%
  dplyr::group_by(sample) %>%
  dplyr::mutate(proportion = norm_count / sum(norm_count)) %>%
  dplyr::filter(!subunit %in% divergent_subunits) %>% # Remove divergent subunits
  dplyr::mutate_if(is.numeric, round, digits = 4)

median_proportions <- amel_bter_brain_proportions %>%
  dplyr::ungroup() %>% # Remove grouping variable previously created
  dplyr::select(species, caste, subunit, proportion) %>%
  dplyr::group_by(species, caste, subunit) %>%
  dplyr::summarise_at(vars(-group_cols()), ~ median(.)) %>%
  dplyr::rename(median_proportion = proportion) %>%
  dplyr::mutate_if(is.numeric, round, digits = 4)

# Verify that the median values match the expected results
amel_bter_brain_proportions %>%
  dplyr::filter(species == "amel" & caste == "queen" & subunit == "alpha1") %>%
  dplyr::pull(var = proportion)
median(c(0.0199, 0.0149, 0.0239, 0.0269))

# Change to wide format
median_proportions_wide <- median_proportions %>%
  tidyr::spread(key = species, value = median_proportion)

queens_median_proportions_wide <- median_proportions_wide %>%
  dplyr::filter(caste == "queen")

workers_median_proportions_wide <- median_proportions_wide %>%
  dplyr::filter(caste == "worker")
```

# Correlation tests between _A mellifera_ and _B terrestris_
Use the median proportion values to perform correlation tests between species.
## Correlation test between queens of _A mellifera_ and _B terrestris_
```{r}
# Run correlation test using the Spearman method
cor.test(
  queens_median_proportions_wide$amel,
  queens_median_proportions_wide$bter,
  method = "spearman"
)

# 	Spearman's rank correlation rho
# 
# data:  queens_median_proportions_wide$amel and queens_median_proportions_wide$bter
# S = 70, p-value = 0.2696
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#       rho 
# 0.4166667 
```

## Correlation test between workers of _A mellifera_ and _B terrestris_
```{r}
# Run correlation test using the Spearman method
cor.test(
  workers_median_proportions_wide$amel,
  workers_median_proportions_wide$bter,
  method = "spearman"
)

# 	Spearman's rank correlation rho
# 
# data:  workers_median_proportions_wide$amel and workers_median_proportions_wide$bter
# S = 90, p-value = 0.5206
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#  rho 
# 0.25 
```